# Industrial Process and Product Use Sector (IPPU) {#sec-ippu}



```{r}
require(tidyverse)
require(sf)
require(patchwork)
require(plotly)
require(magrittr)
require(gt)
require(ggalluvial)
```


```{r}
## Data
ghg = read_csv('../estimates/processed/ghg_master.csv')

# ghg |> distinct(ghg)

ghg = ghg |> 
  ## note that the hfc is converted already
  # mutate(co2equ = case_when(
  #   ghg == "co2" ~ data * 1,
  #   ghg == "ch4" ~ data * 28,
  #   ghg == "n2o" ~ data * 273,
  #   ghg == "hfc" ~ data * 12
  #   ), 
  #   .after = data) |> 
  mutate(co2equ = data , .after = data) |> 
  filter(ghg %in% c('co2', 'hfc') & sector == "IPPU")

ippu_major = ghg |> 
    filter(sector == "IPPU") |> 
    mutate(new_category = category, .after = category) |> 
    separate(new_category, into = c('code', "subcategories"), sep = "-") |> 
    mutate(nn = nchar(code), .after = code)

```




##  Summary of IPPU Sector Activities

This chapter presents the results of the greenhouse gas (GHG) emissions from Industrial Processes and Product Use (IPPU) sector, covering the period from 1990 to 2021. IPPU contributes to the country’s total greenhouse gas emissions from CO~2~, CH~4~, N~2~O and fluorinated gases (F-gases) which are mainly from industrial processes that do not involve direct fuel combustion (@tbl-ippu). Emissions are reported in CO~2~ equivalents (Gg) and analyzed across key categories. 

```{r}
#| label: tbl-ippu
#| tbl-cap: Industrial Processes and Product Use major categories and the greenhouse gases


# ghg |> 
ippu_major |> 
    filter(nn == 4 ) |> 
    # distinct(category, ghg) |> 
    group_by(category, ghg) |> 
    summarise(data = sum(co2equ, na.rm = TRUE)) |> 
    ungroup() |> 
    mutate(state= if_else(data == 0, NA, "YES")) |> 
    select(-data) |> 
    pivot_wider(values_from = state, names_from = ghg)  |> 
    gt() |> 
    gt::tab_spanner(columns = 2:3, label = "Greenhouse Gases") |> 
    tab_options(data_row.padding = px(5)) |> 
    cols_width(category ~ px(500)) |> 
    cols_label(
      category ~'IPPU Category',
      co2 ~ "{{CO_2}}",
      hfc ~ 'HFCs'
    ) |> 
    fmt_missing(columns = 2:3, missing_text = "-") |>
    tab_footnote(
      footnote = md("Activity data to estimate CO~2~ and HFCs were available."),
      locations = cells_column_labels(columns = 2:3), 
      placement = 'left'
    )
```

Emission trends for several categories within the Industrial Processes and Product Use (IPPU) sector were assessed, including 2.A - Mineral Industry, 2.C - Metal Industry, 2.D - Non-Energy Products from Fuels and Solvent Use (6), 2.F - Product Uses as Substitutes for Ozone-Depleting Substances, and 2.H - Other Categories. However, due to insufficient data availability, emission trends for 2.C - Metal Industry, 2.E - Electronics Industry, and 2.G - Other Product Manufacture could not be evaluated (see @tbl-ippu-nonplots).

```{r}
#| label: tbl-ippu-nonplots
#| tbl-cap: Trends of CO~2~eq emission by major categories in Industrial Processes and Product Use 


ippu_major |> 
    filter(nn == 4 ) |> 
    group_by(category, years) |> 
    summarise(co2equ = sum(co2equ, na.rm = TRUE)) |> 
    ungroup() |> 
    pivot_wider(names_from = years, values_from = co2equ) |> 
    janitor::clean_names() |> 
    gt(rowname_col = 'category') |> 
    fmt_number(columns = everything(), decimals = 1) |>     
    cols_nanoplot(
      columns = starts_with("x"), 
      plot_type = 'line',
      new_col_name = "nanoplots",
      new_col_label = md("Trends CO~2~ equ")
    ) |>
    cols_width(nanoplots ~ px(120)) |> 
    cols_align(align = "left", columns = nanoplots) |>
    tab_footnote(
      footnote = "Emission from 1990 through 2021.",
      locations = cells_column_labels(columns = nanoplots)
    )




  
```


##	Methods, data sources and assumptions

### Cement Production
Cement manufacturing is one of the largest contributors to CO₂ emissions in Tanzania's IPPU sector. The production process in cement releases a large amount of CO2 due to the decomposition of limestone (calcium carbonate) to form clinker. Tanzania's cement industry is growing to meet rising domestic and regional demand which results in increasing GHG emissions. However, some producers are exploring efficiency improvements and alternative fuels to reduce emissions. Tanzania’s cement industry has grown significantly, supported by both local and international investments. The industry is led by several key players:

-	**Tanzania Portland Cement Company (TPCC)** - Operating under the brand name **Twiga Cement** is one of the largest cement manufacturers in Tanzania. TPCC operates a major plant at Wazo Hill, Dar es Salaam and has a significant market share, producing around 1.9 million tons annually. The company is part of HeidelbergCement Group.
-	**Dangote Cement** - A Nigerian company, Dangote entered the Tanzanian market in 2015, establishing a $500 million cement plant in Mtwara. This plant, with an annual capacity of 3 million tons is powered by a 30-MW gas-fired power plant and has access to significant limestone reserves. Dangote’s entry increased competition and lowered cement prices which has significantly impacted the industry.
-	**Tanga Cement** - Known for its Simba brand, Tanga Cement operates in the northeastern region and expanded its capacity to 1.25 million tons per year by adding a second production line in 2016. Despite competition pressures, Tanga remains a key player in the market.

- Other notable contributors include subsidiaries of international groups like LafargeHolcim and AfriSam. 

These companies collectively contribute to an installed cement production capacity of approximately 7 million tons annually in Tanzania thus positioning the country as a leading cement producer in East Africa. However, The industry has faced various challenges including the shortage of fuel and electricity which have impacted and reduced production costs and profit margins. 

> The demand for cement remains robust mainly driven by infrastructure and construction projects throughout the country. 


###  Lime Production
Lime production contributes CO~2~ emissions through the calcination process. Although it does not contribute CO~2~ as large as cement, the lime industry is growing with uses in construction, water treatment and agriculture. Emissions from lime production are less significant than from cement but still contribute to the IPPU sector’s GHG output.
Lime production in Tanzania is significant because it is  largely driven by small-scale and traditional methods, often involving coral limestone as the raw material. The lime industry in Tanzania includes both artisanal operations and more structured small to medium enterprises. Key areas for lime production include regions along the coast particularly in places like Mtwara where lime is produced from coral mining. This traditional method involves burning coral boulders collected from shallow reefs, especially in the Mnazi Bay-Ruvuma Estuary Marine Park region which has led to environmental concerns such as reef degradation and coral depletion.

Artisans use live coral from areas like Shangani and Msangamkuu highlighting the reliance on natural marine resources in coastal lime production. Due to its accessibility and quality, coral limestone remains a popular choice for producing "*white lime*" or "chokaa," primarily used in building and agricultural sectors. While some improvements are underway in methods and environmental management, lime production still faces challenges. Many producers rely on rudimentary kilns with limited capacity therefore affecting production efficiency and environmental sustainability. Initiatives to regulate coral mining, enforce sustainable practices and introduce modern lime kilns have been recommended to balance production needs with environmental conservation in Tanzania's lime industry.

### Metal Production and Mining
Iron and steel production in Tanzania is limited but small-scale metal production and mining contribute to GHG emissions through fuel combustion and chemical reactions in the processes. As Tanzania seeks to grow its industrial base, emissions from metal production are likely to rise unless cleaner technologies are adopted. The metal production sector in Tanzania is focused on various metals including iron, steel, nickel and gemstones with significant industrial and investment developments in recent years. The country possesses notable iron ore reserves primarily at the Liganga site in Njombe Region. 

The Liganga Iron and Steel Complex is an extensive USD 1.8 billion investment by Tanzania China International Mineral Resources Ltd (TCI MRL) which aims to produce up to one million tonnes of iron and steel annually along with vanadium pentoxide and titanium dioxide. This facility is designed to not only meet local demand but also export products therefore diversifying Tanzania’s economy. Tanzania’s nickel resources are primarily located in the Kabanga Nickel Deposit in the northwest, one of the world's largest undeveloped nickel sulfide deposits with an estimated 58 million tonnes. Efforts are underway to develop this deposit to support global demand especially as nickel is critical for battery production in electric vehicles. 

The Kabanga Nickel Project is expected to generate a significant amount of nickel with plans to reach production levels that could place Tanzania among the top producers worldwide. Additionally, Tanzania is known for its unique gemstone, tanzanite which is exclusively found in the Mererani Hills in the Manyara Region. This gemstone along with diamonds from the Williamson mine near Mwanza contributes significantly to Tanzania’s mining revenues. Gold production is also substantial, with Tanzania ranking as one of Africa’s leading gold producers.

### Manufacturing of Other Mineral Products
In addition to cement and lime, other mineral products including glass and ceramics also release CO₂ emissions during production. The expansion of these industries to meet local and regional demands is contributing to emissions though to a lesser extent than cement. In Tanzania, the glass and ceramics sector are experiencing notable expansion with several significant investments aimed at meeting both domestic and regional demand. One major project is the Sapphire Float Glass plant in Mkuranga which is expected to start production soon. This $311 million investment is expected to produce up to 1,200 tonnes of float glass daily therefore making it the first of its kind in the country. Located in the Kusini Gateway Industrial Park, the plant aims to reduce Tanzania's dependence on imported glass products and support the growing construction industry. Another notable player is Keda Industrial Group from China which announced an $87 million investment in an architectural glass plant in Tanzania. This facility will produce approximately 600 tonnes of glass annually, focusing on materials used in building windows and external wall applications.

### Use of Hydrofluorocarbons (HFCs) in Refrigeration and Air Conditioning
HFCs are potent greenhouse gases used as refrigerants in air conditioning, refrigeration and other cooling systems. In Tanzania, hydrofluorocarbons (HFCs) are primarily used in refrigeration, air conditioning, foam production and fire suppression. The main types of HFCs include HFC-134a, HFC-32, HFC-125, and HFC-143a, with HFC-134a and HFC-32 being the most used in Tanzania’s growing air conditioning and refrigeration sector. As these sectors expand to meet the needs of urbanization and rising living standards, HFC demand is expected to increase despite global pressures to phase down HFC usage.

Projections suggest that demand for HFCs in Tanzania will grow in the short term but with international agreements encouraging a shift to low-global-warming-potential (GWP) alternatives, the country may start transitioning towards less harmful refrigerants. Currently, Tanzania relies on imports for HFCs primarily sourcing them from Asian countries with China being a major supplier. The Tanzanian government is aware of the need to mitigate HFC emissions and is working on policies to align with the Kigali Amendment’s goals, which mandates gradual reductions in HFC use. By 2040, it is anticipated that Tanzania’s reliance on high-GWP HFCs will significantly decrease as alternatives become more accessible and policy measures are implemented to encourage the shift to environmentally friendly options.

### Other Product Use
Other products such as aerosols, fire extinguishers, and foam-blowing agents also contain greenhouse gases and contribute to Tanzania's IPPU sector emissions. These sources are relatively small but contribute to the overall emissions profile.


##	Overview of Emissions for the IPPU sector 

The IPPU sector has experienced significant growth in GHG emissions over the past two decades, with the most dramatic increases observed in the Mineral Industry (2.A) and Product Uses as Substitutes for Ozone-Depleting Substances (2.F).  In 2021, emissions from the IPPU sector were equivalent to 304,337 Gg-CO~2~eq, an increase of 78,405 Gg-CO~2~eq (288 percent) compared to 2000 (@tbl-ippu-cat). The Mineral Industry (2.A) experienced a notable rise in emissions, escalating from 343 Gg in 2000 to 2,716 Gg in 2021, which translates to a 693% increase. In a similar vein, the Metal Industry (2.C) saw a 300% uptick, with emissions climbing from 99 Gg to 397 Gg (@tbl-ippu-cat). 

The Non-Energy Products from Fuels and Solvent Use (2.D) sector recorded a more modest elevation, advancing from 101 Gg to 161 Gg, indicating a 60% change (@tbl-ippu-cat). The most pronounced growth was observed in the Product Uses as Substitutes for Ozone-Depleting Substances (2.F), where emissions skyrocketed from 44,430 Gg to 191,256 Gg, reflecting a 330% increase (@tbl-ippu-cat). This category makes the largest absolute contribution to emissions within the IPPU sectors. Finally, the Other (2.H) category also marked a significant increase, with emissions rising from 33,433 Gg to 109,808 Gg, representing a 228% change (@tbl-ippu-cat).



```{r}
#| label: tbl-ippu-cat
#| tbl-cap: Emissions from Industrial Processes and Product Use (IPPU) by categories

 
summary_all = ippu_major |> 
    filter(nn == 4 & data > 0 & years %in% c(2000, 2021)) |> 
    group_by(years) |> 
    summarise(total = sum(co2equ, na.rm = TRUE)) |> 
    ungroup() |> 
    pivot_wider(values_from = total, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021 - x2000) / x2000*100 ) |> 
    mutate(category = "TOTAL", .before = x2000)

summary_cat = ippu_major |> 
    filter(nn == 4 & data > 0 & years %in% c(2000, 2021)) |> 
    group_by(years, category) |> 
    summarise(total = sum(co2equ, na.rm = TRUE)) |> 
    ungroup() |> 
    pivot_wider(values_from = total, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021 - x2000) / x2000*100 ) 

summary_cat |> 
    bind_rows(summary_all) |> 
    gt() |> 
    fmt_number(columns = everything(), decimals = 0) |> 
    tab_spanner(columns = 2:3, label = md("CO~2~eq (Gg)")) |> 
    cols_label(
        category = "IPPU categories",
        x2000 = '2000',
        x2021 = '2021',
        pchange = 'Change (%)'
    ) |>     
    tab_options(
        table.width = px(700), 
        column_labels.font.weight = "bold"
    )  |>     
  tab_style(
    style = cell_text(weight = "bold"),  
    locations = cells_body(rows = 6)    
  )|>
  cols_add(dir = ifelse(x2021 > x2000, "red", "forestgreen")) |>
  cols_label(dir = "") |>
  text_case_match(
    "red" ~ fontawesome::fa("arrow-up"),
    "forestgreen" ~ fontawesome::fa("arrow-down")
  )|>
  tab_style(
    style = cell_text(color = from_column("dir")),
    locations = cells_body(columns = dir)
  )|> 
  # specify width of columns
  cols_width(
    category ~ px(300),  
    x2000 ~ px(70),    
    x2021 ~ px(70),    
    pchange ~ px(70),
    dir ~ px(70)
  )


```


##    2.A - Mineral Industry 
The Mineral Industry is a major source of emissions due to its reliance on energy-intensive processes and carbonate-based raw materials [@wang2024accelerated; @segal2024critical]. This section presents an overview of emissions from the Mineral Industry, focusing on key processes and activities that contribute to greenhouse gas emissions (@tbl-mining). The main categories include cement production [@griffiths2023decarbonizing], which generates significant CO~2~ emissions due to the chemical breakdown of limestone and the burning of fossil fuels; lime production, where emissions result from heating limestone to high temperatures; and glass production, which releases emissions during the melting of raw materials like silica and soda ash [@segal2024critical]. The category other uses of soda ash, which is commonly used in detergents and chemicals; and non-metallurgical magnesia production, where magnesia is produced from magnesium carbonate, releasing CO~2~ in the process. 

```{r}
mineral_categories = categories <- c(
    "2.A.1 - Cement production",  "2.A.2 - Lime production",  "2.A.3 - Glass Production",
  "2.A.4 - Other Process Uses of Carbonates",  "2.A.4.a - Ceramics",  "2.A.4.b - Other Uses of Soda Ash",  "2.A.4.c - Non Metallurgical Magnesia Production"
  )

mining = ghg |> 
  filter(category %in% mineral_categories )


```

```{r}
#| label: tbl-mining
#| tbl-cap: subcategories in mineral industry category of the IPPU
#| 
mining |> 
  group_by(category, years) |> 
  summarise(total = sum(co2equ, na.rm = TRUE)) |> 
  ungroup() |> 
  pivot_wider(names_from = years, values_from = total) |> 
  janitor::clean_names() |> 
  gt(rowname_col = 'category') |> 
  cols_nanoplot(
    columns = starts_with("x"), 
    plot_type = 'bar', 
    missing_vals = 'marker', 
    autoscale = FALSE,
    new_col_name = "nanoplots",
    new_col_label = md("Trends CO~2~ equ")
    ) |>
    cols_width(nanoplots ~ px(120)) |> 
    cols_align(align = "left", columns = nanoplots) |>
    tab_footnote(
      footnote = "Emission from 1990 through 2021.",
      locations = cells_column_labels(columns = nanoplots)
    )
```


### 2.A.1 - Cement production

Emissions from cement production steadily increased with time from the 1990s (@fig-cement).  In 2021, CO~2~ from cement production was 2,580Gg, an increas of around seven folds (694%) compared to 325Gg emitted in 2000 (@fig-cement). The increasing trend of emission from cement production is attributted with major growth occurred in the 2000s as new plants opened and existing ones expanded capacity to meet demand. By the mid-2010s, Tanzania became a regional cement exporter. Production reached over 6 million tons by 2020, supported by local and foreign investment. 


```{r}
#| label: fig-cement
#| fig-cap: CO~2~ equivalent emission from  Cement production
#| fig-width: 8
#| fig-height: 4


dt = mining |> 
  filter(category %in% "2.A.1 - Cement production")

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 1)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 1.8e3,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 2.8e3), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


### 2.A.2 - Lime production
Lime production in Tanzania grew moderately from the 1990s through 2009, largely due to demand from construction, agriculture, and mining sectors. The industry remained mostly small-scale and informal in the earlier years, with production concentrated in a few regions. As infrastructure projects expanded in the 2000s, lime production rose, prompting some modernization efforts. By 2020, production was influenced by increased cement production and industrial demand. However, the sector continued facing challenges, including inconsistent quality standards and limited access to advanced production technology.




```{r}
#| label: fig-lime
#| fig-cap: CO~2~ equivalent emission from lime production
#| fig-width: 8
#| fig-height: 4


dt = mining |> 
  filter(category %in% mineral_categories[2])

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 1)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = .8e2,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 1.1e2), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


### 2.A.3 - Glass Production

The emissions generated from glass production increased from 1994 to2010, then later decreased in 2011 and rose again at a moderate rate until 2021. The yearly variations in glass production in Tanzania can be influenced by several factors. Fluctuations in demand from construction and packaging industries can lead to production adjustments. Additionally, changes in raw material availability and costs, particularly silica sand, can impact output levels. Economic conditions, including inflation and currency exchange rates, also affect production costs. Lastly, regulatory changes and environmental policies may impose constraints or incentives that influence production capabilities.


```{r}
#| label: fig-glass
#| fig-cap: CO~2~ equivalent emission from glass production
#| fig-width: 8
#| fig-height: 4


dt = mining |> 
  filter(category %in% mineral_categories[3])

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>")
    #   x2000 = scales::comma(x = x2000, big.mark = ","),
    #   x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 3.2e0,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 4.3e0), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


###	Category 2A4 - Other Process Uses of Carbonates

This category emission of CO~2~ equivalent is interely from Uses of Soda Ash (@tbl-mining), which has been influenced by demand from industries like glass manufacturing, detergents, and mining. Since Tanzania does not produce soda ash domestically, the use of imported Soda Ash have been the main source emission (@tbl-mining). 


In 2021 the use of Soda Ash alone in Other Process Uses of Carbonates category emitted about 43.8Gg of CO~2~, which is more than two folds (215.4%) rise compared to 13.9Gg emitted in 2000 (@fig-2a4). The sharp increase in CO~2~ emission from 2013 (*See* subplot @fig-2a4) is mainly contributed with exponential increase of Soda Ash  (@tbl-mining). The  increase of CO~2~ from 1990 to 2012 relatively steady as there was no data during this period and were refilled through interpolation (@fig-2a4). 


```{r}
#| label: fig-2a4
#| fig-cap: CO~2~ equivalent emission from  Other Process Uses of Carbonates
#| fig-width: 8
#| fig-height: 4


dt =  mining |> 
  filter(category %in% mineral_categories[4])
  

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 


subplot =  dt |> 
  filter(co2equ > 0 & years > 2012) |> 
  ggplot(aes(x = years, y = co2equ)) +  
  geom_line(color = "steelblue4",  linewidth = .8) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 4.6e1), expand = c(0,NA), name = expression(CO[2]~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,2))+
  theme_minimal(base_size = 11)+
  theme(panel.background = element_rect(fill = scales::alpha(colour = 'darkblue', alpha = 0.1)), axis.title.x = element_blank(), panel.grid.minor = element_blank(), panel.grid.major = element_line(linetype = 2, linewidth = .1))+
  labs(subtitle = "Emission: 2012-2021")


  aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>")
    #   x2000 = scales::comma(x = x2000, big.mark = ","),
    #   x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 3.6e1,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )


 
 mainplot = dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ)) + 
    annotate(geom = "rect", xmin = 1989, xmax = 2013, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
    ) + 
    annotate(geom = "rect", xmin = 2011, xmax = 2020, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
    )  +
    annotate(geom = "rect", xmin = 2018, xmax = 2022, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
    )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 4.6e1), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  +
        annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5) 


 mainplot + inset_element(p = subplot,  left = 0.5, bottom = 0.55, right = .8, top = .95, ignore_tag = TRUE)

```





###	Category 2D1 & 2D2 - Lubricant and paraffin wax use

The emission trend of lubricant use followed the stable use patterns until 2013, where it rapidly  decreased in 2014 and thereafter lubricant use was observed to increase from 2015 onwards. Since 2015, rising vehicle numbers and industrial machinery have bolstered demand, with a notable rise in imports during the mid-2000s as these sectors expanded. Economic fluctuations and oil price changes affected volumes, but demand increased due to industrial needs.

Emissions from paraffin wax, generally increasing with time. This was attributed to paraffin wax imports in Tanzania which have generally followed demand from sectors such as candle making, packaging, and cosmetics. Since the 1990s, imports have shown gradual growth, with peaks during periods of industrial and artisanal production expansion. Demand for paraffin wax increased in the early 2000s as small and medium enterprises, especially in candle production, grew. By 2020, imports remained essential.


```{r}
#| label: fig-lubricants
#| fig-cap: CO~2~ equivalent emission from  Lubricant and paraffin wax use
#| fig-width: 8
#| fig-height: 4


dt = ghg |> 
  filter(category %in% c('2.D.1 - Lubricant Use' , '2.D.2 - Paraffin Wax Use'))

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100) |>
    mutate(category = c("Lubricant", "Paraffin"))  
    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 235,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000 |> round(1)}Gg/year<br>
        2021 = {aa[1,]$x2021 |> round(1)}Gg/year<br>
        Change = {aa[1,]$pchange |> round(1)}%"
        )
    )

 lab_df2 = aa[2,] |>    
    mutate(
      years = 2000,  co2equ = 235,
      lab_stats = glue::glue(
        "**{aa[2,]$category |> str_to_upper()}**<br>
          ------<br>
          2000 =  {aa[2,]$x2000 |> round(1)}Gg/year<br>
          2021 = {aa[2,]$x2021 |> round(1)}Gg/year<br>
          Change = {aa[2,]$pchange |> round(1)}%"
          )
      ) 


 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ, fill = category))+
  annotate(geom = "rect", xmin = 1989, xmax = 2007, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2005, xmax = 2020, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2018, xmax = 2024, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  )  +    
  geom_area(color = "ivory") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  +
  ggtext::geom_richtext(
    data = lab_df2, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5,
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )           

```




### Category 2.F.3 - Fire Protection

Emissions from fire protection systems display an abrupt shift (@fig-fire). Negative values from 1990 to 2006, indicating potential data discrepancies, accounting conventions, or net removals during that period. From 2007 onward, a substantial increase is observed, peaking at 6657.4 in 2021. In Tanzania, the import and use of fire protection gases have grown as safety standards improve across commercial, industrial, and public spaces. From the 2000s onward, demand increased particularly in sectors like data centers, manufacturing, and large buildings. In recent years, environmental awareness has influenced a shift towards eco-friendly alternatives, aligning with global standards on fire suppression. By 2020, Tanzania continued to import fire protection gases, with an emphasis on sustainable and effective fire safety solutions compliant with international environmental protocols.



```{r}
#| label: fig-fire
#| fig-cap: CO~2~ equivalent emission from fire protection
#| fig-width: 8
#| fig-height: 4

dt = ghg |> 
  filter(category %in% c('2.F.3 - Fire Protection')) |> 
  drop_na()

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 25),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 5.6e3,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  # filter(co2equ > 0) |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1989, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(-200, 1.2e4), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```

### Category 2.F.4 - Aerosols

In Tanzania, aerosol imports have steadily increased, driven by demand in personal care, household products, and industrial applications. The usage rose with urbanization and the growth of consumer goods products like sprays and deodorants. In recent years, environmental considerations have influenced a shift towards eco-friendly propellants as regulations on ozone-depleting substances have tightened. The significant jump in emissions in 2020 and 2021 is unusual and could result from changes in production methods and increased aerosol use. However, aerosol imports remained vital, with a focus on more sustainable and compliant formulations to meet both consumer demand and environmental standards.



```{r}
#| label: fig-aerosols
#| fig-cap: CO~2~ equivalent emission from Aerosols
#| fig-width: 8
#| fig-height: 4

dt = ghg |> 
  filter(category %in% c('2.F.4 - Aerosols')) |> 
  drop_na()

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 25),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 400,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  # filter(co2equ > 0) |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1989, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(-200, .9e3), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


### Category 2.F.5 - Solvents

Tanzania’s import trends for solvents have generally been tied to demand in industries such as manufacturing, paint production, and pharmaceuticals. This contributed to a general increase in emissions with time (@fig-solvents). Imports of solvents saw gradual growth as these sectors expanded, especially in urban areas. Increased industrialization and the use of volatile organic compounds (VOCs) in manufacturing in the early 2000s spurred more demand, and import volumes continued to rise steadily. 




```{r}
#| label: fig-solvents
#| fig-cap: CO~2~ equivalent emission from Solvents
#| fig-width: 8
#| fig-height: 4

dt = ghg |> 
  filter(category %in% c('2.F.5 - Solvents')) |> 
  drop_na()

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 25),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 4e3,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  # filter(co2equ > 0) |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1989, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(-200, 7.2e3), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


###	Category 2F1 - Refrigeration and Air Conditioning

Generally the emission trend from the refrigerant uses increased consistently with time. Refrigerant imports in Tanzania have generally increased alongside rising demand for air conditioning and refrigeration, driven by economic growth and rising temperatures. From the 1990s, imports rose steadily as commercial and residential cooling needs expanded. In the 2010s, efforts to phase out ozone-depleting refrigerants aligned with global environmental agreements, leading to a gradual shift toward more eco-friendly refrigerants. By 2020, Tanzania’s imports continued to support cooling needs, though there was growing attention to sustainable alternatives and compliance with environmental standards.

```{r}
#| label: fig-refigeration
#| fig-cap: CO~2~ equivalent emission from  Refrigeration and Air Conditioning
#| fig-width: 8
#| fig-height: 4

dt = ghg |> 
  filter(category %in% c('2.F.1.a - Refrigeration and Stationary Air Conditioning'))

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 25),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 3.6e3,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1989, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 5.6e3), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```



###	Category 2G1 - Electrical Equipment



### Iron and Steel Production

The emissions from iron and steel production generally increased with time (@fig-iron) showing a significant upward trend in emissions. This growth reflects an increased production capacity and demand for iron and steel over the years, driven by industrialization and infrastructure development. Rising infrastructure projects and economic growth often demand more steel for construction, machinery, and transportation. Developing countries like Tanzania may experience rapid growth in this sector due to industrial expansion.

```{r}
#| label: fig-iron
#| fig-cap: CO~2~ equivalent emission from Iron and Steel Production
#| fig-width: 8
#| fig-height: 4


dt = ghg |> 
  filter(category %in% '2.C.1 - Iron and Steel Production')

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000,2010, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>")
    #   x2000 = scales::comma(x = x2000, big.mark = ","),
    #   x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 3.2e2,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 4.3e2), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


## Others

###	Category 2H1 - Pulp and Paper Industry

Generally the trend of emissions generated from pulp and paper production was increasing with time as pulp and paper production has been driven primarily by demand for packaging, newspapers, and office supplies. From the 1990s, production grew slowly, supported by government and private investments, though constrained by limited raw material availability and outdated machinery. The early 2000s saw slight improvements as Tanzania aimed for self-sufficiency in paper production. By 2020, environmental concerns and competition from imported products posed challenges, but the growing demand for packaging materials and sustainable paper alternatives continued to support the industry’s relevance.

```{r}
#| label: fig-pulp
#| fig-cap: CO~2~ equivalent emission from  Pulp and Paper Industry
#| fig-width: 8
#| fig-height: 4


dt = ghg |> 
  filter(category %in% c('2.H.1 - Pulp and Paper Industry'))

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100) |>
    mutate(category = c("Pulp and Paper<br> Industry"))  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 1.146e5,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000 |> round(1)}Gg/year<br>
        2021 = {aa[1,]$x2021 |> round(1)}Gg/year<br>
        Change = {aa[1,]$pchange |> round(1)}%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1989, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6),labels = scales::label_number(big.mark = ','), limits = c(0, 1.46e5), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```

###	Category 2H2 - Food and Beverages Industry

Food and beverages industry included emissions from biscuit and cake production, sugar production, fish production, animal feed production, bottled beer production, chibuku production, spirits production and coffee beans production 

```{r}
#| label: fig-food
#| fig-cap: CO~2~ equivalent emission from  Food and Beverages Industry
#| fig-width: 8
#| fig-height: 4


dt = ghg |> 
  filter(category %in% c('2.H.2 - Food and Beverages Industry'))

# dt |>drop_na() |>filter(years %in% c(2000,2021)) 

aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100) |>
    mutate(category = c("Food and Beverages<br> Industry"))  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 4.3e3,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000 |> round(1)}Gg/year<br>
        2021 = {aa[1,]$x2021 |> round(1)}Gg/year<br>
        Change = {aa[1,]$pchange |> round(1)}%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1989, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  # gghighlight::gghighlight(max(co2equ) > 0)+
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6),labels = scales::label_number(big.mark = ','), limits = c(0, 5.4e3), expand = c(0,NA), name = expression(CO[2]~equivalnent~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  # scale_fill_manual(values = c('2.D.1 - Lubricant Use'='cyan4', '2.D.2 - Paraffin Wax Use'='steelblue4'))+
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


## Other food

This section provides an overview of CO~₂~ emissions associated with the production of various food and beverage categories over the inventory years. @tbl-otherfoods highlights significant trends and variations across different products, reflecting changes in production levels, energy use, and industrial practices. While some categories, such as sugar and spirits, show substantial increases in emissions due to their energy-intensive processes, others, like fish and Chibuku, show more stable trends. Notably, biscuits and bottle production have seen dramatic growth in emissions in recent years, underscoring the environmental impact of expanding production scales. These trends  and descriptive statistics are further described in  @sec-feed through @sec-spirit.

```{r}
others = readxl::read_excel(
  path = "../estimates/IPPU Sector Summary Emmisions.xlsx", 
  sheet = "Other (Food and Bev)") |>
  rename(years=1) |> 
  mutate(years = as.integer(years))

others_names = others |> names()

others_long = others |> 
  pivot_longer(cols = -1, values_to = 'co2equ', names_to = 'category')

```


```{r}
#| eval: false
#| 
others_long |> 
   pivot_wider(names_from = years, values_from = co2equ) |> 
   mutate(across(is.numeric,round, 1)) |> 
   gt(rowname_col = 'category') |> 
   cols_nanoplot(columns = everything(), plot_type = 'bar')
```



```{r}
#| label: tbl-otherfoods
#| tbl-cap: Heatmap of CO~2~ emission from  production of various food and beverage categories
others_long |> 
  # filter(category %in% c('Bottled beer production', 'Spirits production', 'Sugar Production')) |> 
  pivot_wider(names_from = years, values_from = co2equ) |> 
  mutate(across(is.numeric,round, 1))  |> 
  mutate(category = c("Feed", "Biscuits", "Bottle", "Chibuku", "Fish", "Spirits", "Sugar")) |> 
  arrange(`2021`) |> 
  select(c(1,seq(2,25,10),27:33)) |> 
  gt() |> 
  tab_spanner(columns = 2:8, label = "Emission from Food Production (CO2e (Gg))") |> 
  cols_width(category ~ px(400)) |> 
  data_color(
    columns = 2:11,
    method = "numeric",
    direction = "row",
    palette = hcl.colors(n = 20, palette = 'Earth', rev = TRUE)
    )
  # hcl.pals(type = 'sequential')
  # hcl.pals(type = 'divergingx')
```


###	Animal feed production {#sec-feed}

Emissions from animal feed production started at 2.579 in 1990, showing a gradual increase over the years, with a significant spike in 2021 (13.781). The trend of emissions from animal feed production has been increasing due to the growth of the livestock sector and rising demand for poultry and dairy products. Local production facilities have expanded, focusing on improving quality and accessibility. However, imports remain necessary, particularly for specialized feeds and raw materials not produced domestically. By 2020, the trend showed a combination of local production efforts to meet the growing needs of the agricultural sector, balancing between self-sufficiency and reliance on foreign sources.


```{r}
#| eval: false


 others_long |> 
  ggplot(aes(x = years, y = co2equ, fill = category))+
  geom_area()+
  ggsci::scale_fill_lancet()+
  theme(legend.position = c(.3,.8)) +
  guides(fill = guide_legend(nrow = 4))
```




```{r}
#| label: fig-feed
#| fig-cap: CO~2~ equivalent emission from Animal feed production
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[2])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>")
    #   x2000 = scales::comma(x = x2000, big.mark = ","),
    #   x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 1.0e1,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 0.14e2), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


###	Biscuit and cake production {#sec-biscuits}

Biscuit and cake production has grown alongside demand from urban populations and the expanding retail sector. Since the 2000s, local production increased as both small bakeries and larger manufacturers scaled operations to meet consumer demand for affordable snacks. However, imports of biscuits and premium cakes continue to fill gaps in supply, especially for specialty products. By 2020, production remained strong, with a focus on local sourcing and variety, while imports supplemented market needs for diverse and premium options.



```{r}
#| label: fig-biscuts
#| fig-cap: CO~2~ equivalent emission from Biscuits and cakes production
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[3])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>")
    #   x2000 = scales::comma(x = x2000, big.mark = ","),
    #   x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 2.0e1,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 3.14e1), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


###	Fish products {#sec-fish}
Emissions increased from 9.2649 in 1990 to 13.7598 in 2017, showing slight fluctuations in recent years. The fishing industry has grown steadily, driven by domestic and export demand for fish products. Peaks and troughs in emissions may correspond to variations in fish harvests, regulatory changes, or shifts in processing methods. The decline in fish production is often due to overfishing, which depletes fish populations and affects sustainability. Environmental degradation, such as pollution and habitat loss, particularly in breeding areas, further reduces fish stocks. Climate change also impacts fish habitats by altering water temperatures and conditions, disrupting breeding cycles. Additionally, inadequate management and enforcement contribute to unregulated fishing, while limited investment in aquaculture keeps the industry heavily reliant on natural stocks. Together, these factors pose significant challenges to maintaining fish production levels.


```{r}
#| label: fig-fish
#| fig-cap: CO~2~ equivalent emission from fish products
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[6])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>")
    #   x2000 = scales::comma(x = x2000, big.mark = ","),
    #   x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 1.60e1,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 2.14e1), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```

 

###	Sugar production {#sec-sugar}
Sugar is a staple product with high demand in both domestic and export markets. Sugar production in Tanzania has grown steadily over the years, though it hasn’t fully met domestic demand. Local production is primarily from sugarcane plantations, but factors like outdated technology and weather variability limit output. Industrial expansion and improved production capacities have contributed to higher emissions. The significant growth reflects the sector's importance to Tanzania’s economy, with emissions likely driven by increased use of fossil fuels in production.

 
```{r}
#| label: fig-sugar
#| fig-cap: CO~2~ equivalent emission from sugar production
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[8])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 3.60e3,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5),labels = scales::label_number(big.mark = ','), limits = c(0, 5.14e3), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```


###	Bottled beer production {#sec-bottle}
Emissions have consistently risen, starting at 16.6 in 1990 and peaking at 144 in 2018. A slight decline follows, stabilizing around 133 in 2021. The growth of Tanzania’s beverage industry, along with increased beer consumption driven by urbanization and changing consumer lifestyles, contributed to this trend. Declines post-2018 could result from market saturation, policy changes, or shifts toward alternative beverages.
 
```{r}
#| label: fig-bottle
#| fig-cap: CO~2~ equivalent emission from bottled beer production
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[4])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 1.10e2,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6),labels = scales::label_number(big.mark = ','), limits = c(0, 1.5e2), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```

###	Chibuku production {sec-chibuku}
Generally, emissions from chibuku production show a steady increase from 4.45 in 1990 to a peak of 9.28 in 2016, followed by fluctuations. The observed decline in chibuku production can be attributed to several factors including, an increase in competition from commercial beers and other alcoholic beverages, changing consumer preferences towards more diverse and premium options, and potential regulatory challenges. Additionally, economic factors such as rising production costs and changes in taxation may have impacted profitability. As a result, chibuku, traditionally a popular sorghum beer, faces challenges in maintaining its market share amid a shifting beverage landscape. Chibuku, a traditional African beer, remains popular in rural areas. The moderate growth trend reflects its niche market compared to bottled beer. Emissions from chibuku production are relatively stable compared to bottled beer, possibly due to limited production scalability.


 
```{r}
#| label: fig-chibuku
#| fig-cap: CO~2~ equivalent emission from chibuku production
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[5])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 0.90e1,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6),labels = scales::label_number(big.mark = ','), limits = c(0, 1.2e1), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```

###	Spirits production {#sec-spirit}
Increased demand for spirits, particularly in urban areas, has fueled the expansion of production facilities and outputs. The rapid increase post-2000 aligns with broader trends in beverage industrialization and consumption patterns. The observed increase in the emissions trend generated from spirits production is attributed to an experienced growth in spirits production due to rising consumer demand and the emergence of local distilleries. The industry has seen a mix of traditional and modern production methods, reflecting both cultural practices and changing preferences. While local production has expanded, imports of premium spirits have also increased, catering to a more affluent market.



 
```{r}
#| label: fig-spirit
#| fig-cap: CO~2~ equivalent emission from spirit production
#| fig-width: 8
#| fig-height: 4


dt = others_long |> 
  filter(category %in% others_names[7])


aa =  dt |> 
    filter(years %in% c(2000, 2021)) |> 
    drop_na() |> 
    select(category, years, co2equ) |> 
    pivot_wider(values_from = co2equ, names_from = years) |> 
    janitor::clean_names() |> 
    mutate(pchange = (x2021-x2000)/x2000*100 ) |>
    mutate(across(is.numeric, round, 2)) |> 
    mutate(
      category = dt |> slice(1)  |> pull(category) |> 
        str_wrap(width = 15),category = str_replace_all(category, "\n", "<br>"),
      x2000 = scales::comma(x = x2000, big.mark = ","),
      x2021 = scales::comma(x = x2021, big.mark = ",")     
    )  

    
   lab_df = aa[1,] |>    
    mutate(
      years = 1991,  co2equ = 2.90e2,
      lab_stats = glue::glue(
        "**{aa[1,]$category |> str_to_upper()}**<br>
        ------<br>
        2000 =  {aa[1,]$x2000}Gg/year<br>
        2021 = {aa[1,]$x2021}Gg/year<br>
        Change = {aa[1,]$pchange }%"
        )
    )

 
 dt |> 
  drop_na() |> 
  ggplot(aes(x = years, y = co2equ))+
  annotate(geom = "rect", xmin = 1990, xmax = 2006, ymin = -Inf, ymax = Inf,fill = "cyan4", alpha = 0.2
  ) + 
  annotate(geom = "rect", xmin = 2004, xmax = 2015, ymin = -Inf, ymax = Inf,fill = "steelblue", alpha = 0.2
  )  +
   annotate(geom = "rect", xmin = 2013, xmax = 2021, ymin = -Inf, ymax = Inf,fill = "firebrick", alpha = 0.2
  )  +    
  geom_area(color = "steelblue2", fill = "steelblue4") + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 6),labels = scales::label_number(big.mark = ','), limits = c(0, 5.2e2), expand = c(0,NA), name = expression(CO[2]~e~(Gg)))+
  scale_x_continuous(breaks = seq(1990,2020,5)) +
  theme_minimal(base_size = 11) +
  theme(axis.title.x = element_blank(), legend.position = c(.82,.85), legend.title = element_blank(), panel.grid.minor = element_blank()) +
  # annotate(geom = "label", x = 2000, y = 22, label = "Interpolation\nFlatten Data", alpha = .5, color ='maroon', size = 8.5, lineheight = .8, family = "Barlow Semi Condensed", fontface = "plain", angle = 11.5)   +
  ggtext::geom_richtext(
    data = lab_df, 
    aes(x = years, y = co2equ, label = lab_stats),    hjust = 0, vjust = 0.5, 
    size = 3.5, lineheight = 1.2, family = "Gill Sans MT", fontface = "plain", 
    fill = "#ffffff", color = "black", alpha = .6,
    label.padding = unit(c(0.75, 0.5, 0.5, 0.5), "lines"),
    show.legend = FALSE
   )  

```

## Key categories and recalculations in the IPPU sector 
### Key categories in the IPPU sector  

The IPCC Guidelines provide a framework for identifying and reporting key categories in greenhouse gas (GHG) inventories. Key categories are those that have a significant impact on a country's total emissions or removals and are prioritized for more accurate estimation and reporting [@ipcc2006]. Based on @tbl-key-categories, the primary categories responsible for 95% of emissions in the IPPU sector include: Pulp and Paper Industry (82.4%), Fire Protection (6.3%), Food and Beverages Industry (3.2%), Sugar Production (2.9%), and Refrigeration and Stationary Air Conditioning (2.0%). These categories are the leading sources of emissions. The remaining categories have minimal contributions (≤ 0.2% each) and together account for the last 5% of emissions.


```{r}
#| label: tbl-key-categories
#| tbl-cap: Key categories of activity in emission of CO~2~ in the IPPU Sector
#| 
ghg |> 
  filter(!category %in% c(
  "2 - Industrial Processes and Product Use",
  "2.A - Mineral Industry",
  "2.A.4 - Other Process Uses of Carbonates",
  "2.C - Metal Industry",
  "2.D - Non-Energy Products from Fuels and Solvent Use  (6)",
  "2.F - Product Uses as Substitutes for Ozone Depleting Substances",
  "2.F.1 - Refrigeration and Air Conditioning",
  "2.H - Other")
) |> 
  filter(co2equ > 0) |> 
  select(years, category, co2equ) |> 
  separate(category, into = c('code', 'category'), sep = ' - ') |> 
    select(-code) |> 
  bind_rows(
    others_long
  ) |> 
    group_by(category) |> 
    summarize(total = sum(co2equ, na.rm = TRUE)) |> 
    mutate(pct = total/sum(total) * 100) |> 
    arrange(desc(pct)) |> 
    mutate(cum = cumsum(pct)) |> 
    ungroup() |>     
    gt(rowname_col = 'category') |> 
    fmt_number(columns = everything(), decimals = 1) |> 
    # data_color(columns = 'total', palette = hcl.colors(n = 20, palette = 'Earth'), na_color = NA) |> 
    # data_color(columns = 'pct', palette = 'Reds') |> 
    data_color(columns = 'cum', palette = "Blues", reverse = TRUE, contrast_algo = 'wcag') |> 
    cols_label(
      total = md('CO~2~e(Gg)'),
      pct = md('Contribution'),
      cum = md("Cumulative")
    ) |> 
      tab_spanner(columns = pct:cum, label = "Percentage")
```


### Description of recalculation {#sec-recalculation}

The recalculation of emissions in the **Industrial Processes and Product Use (IPPU) sector** involved a systematic approach to address data gaps and improve the accuracy of greenhouse gas (GHG) emission estimates. The process began with the collection of key data types, including **production volumes**, **import and export data**, and **economic or population indicators**, which are critical for estimating emissions from various industrial activities. These data were used to assess emissions across multiple subcategories within the IPPU sector.

To address missing or incomplete data, three primary techniques were employed: **interpolation**, **extrapolation**, and **surrogate methods**. For six subcategories—**lime production**, **pulp and paper production**, **biscuits and cakes production**, **bottled beer production**, **chibuku production**, and **spirit production**—the **interpolation technique** was applied. This method estimated missing activity data using a linear interpolation equation, as outlined in the IPCC 2006 Guidelines [@ipcc2006]. The mathematical formula used known data points before and after the missing period to estimate the values for the gaps is illustrated in @eq-interp.

$$
Y_t = Y_{start} + \frac{(T_t - T_{start})}{(T_{end}-T_{start})} \times (Y_{end}-Y_{start})
$${#eq-interp}

Whereby: 

> $Y_o$ = missing activity data for relevant time period; $Y_t$ = Missing activity data for relevant time period; $Y_{Start}$ = Activity data value prior to the relevant missing data; 
$Y_{end}$ = Activity data value after the missing data; $T_t$ = Year of the relevant missing data; $T_{start}$ = Year prior to the relevant missing data; $T_{end}$ = Year after the relevant missing data


For subcategories where interpolation was not feasible, the **extrapolation technique** was used. This approach relied on **population growth rates** as a proxy for production trends, based on the assumption that production levels are proportional to population growth. Population data from various sources, including national censuses and projections, were used to establish growth rates, which were then applied to estimate missing activity data. The mathematical formula for extrapolation is presented in @eq-extra

$$
Y_o = \frac{Y_t}{Y_{end}} \times Y_s
$${#eq-extra}


Whereby: 

> $Y_o$ = missing activity data for relevant time period; $Y_t$ = population in year of the relevant missing data; $Y_{end}$ = Population in year after the relevant missing data; and $Y_s$ = Sum of all available activity data 

The **surrogate technique** was applied to six subcategories where related activity data were available. This method used strongly correlated surrogate data to estimate missing values, following the procedures outlined in the IPCC 2006 Guidelines. The mathematical formula for surrogate technique is presented in @eq-surro;

$$
Y_o = Y_t \times \frac{S_o}{S_t}
$${#eq-surro}

Whereby:

> $Y_o$ = missing activity data for relevant time period; $Y_t$ = Activity data after the relevant missing data; $S_o$ = Surrogate data in a year of the relevant missing data; and $S_s$ = Surrogate data in a year after the relevant missing data

These recalculation methods ensured the consistency and completeness of the IPPU sector’s emission inventory, aligning it with the IPCC’s best practices. 


##	Quality assurance/ quality control
General Quality Control (QC)) procedures involved standard quality checks associated with calculations, data processing, completeness, and documentation. These checks applied to all categories of inventory sources and sinks. As specified in the IPCC Guidelines, general inventory-level QC procedures present a set of standard checks that inventory compilers should consistently employ during the inventory preparation process [@ipcc2006]. The checks enumerated in Table 11 should be carried out irrespective of the data type used for developing the inventory estimates. These checks were also pertinent for categories where default values or national data were used as the foundation for the estimates.

##	Description of uncertainties
##	Time series consistency issues
##	Improvements
The Industrial Processes and Product Use (IPPU) sector is a significant contributor to greenhouse gas (GHG) emissions in Tanzania due to its reliance on energy-intensive processes and high-emission product manufacturing, such as cement, steel and chemicals. To mitigate these challenges and align with international climate commitments, several improvements are planned to modernize and decarbonize the sector.

One key area of focus is technological upgrades. Industries are encouraged to adopt energy-efficient innovations such as waste heat recovery systems in cement and steel production in order to reduce energy consumption and emissions intensity. Transitioning to low-carbon technologies like carbon capture and storage for processes such as clinker production is also being prioritized. Additionally, the use of alternative raw materials such as fly ash and slag in cement production or recycled steel can significantly minimize the environmental footprint of industrial activities. 

The transition to renewable energy is a cornerstone of the sector’s decarbonization strategy. Integrating solar, wind and hydropower into industrial processes will reduce reliance on fossil fuels. Investments in energy storage solutions will ensure a reliable energy supply for industries adopting renewable energy sources. Enhanced data collection systems will provide accurate emissions data while the planned periodic reviews shall allow for updates to strategies based on technological advancements and international climate goals. These planned improvements are expected to reduce GHG emissions significantly therefore contributing to Tanzania's Nationally Determined Contributions (NDCs) under the Paris Agreement.





